---
title: "Danish Kings"
author: "Konrad Voldby"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```



The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: ,   

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# Look at line 31-40 in your script and make sure this is at the top
library(tidyverse)
# FILL IN THE CODE BELOW and review the outputs
kings1 <- read.csv("data/kings.csv")

kings2 <- read_csv("data/kings.csv")

kings3 <- read.csv2("data/kings.csv")

kings4 <- read_csv2( "data/kings.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
  Answer: read_csv is the tidverse function that i can use to read my data into a usable object. 
  
2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
  Answer: The class of the kings object created with read_csv is a tibble, which is a modern version of a data frame that provides better printing and subsetting capabilities.

3. How many columns does the object have when created with these different functions?
  Answer: the number of collums in kings1 and kings2 is 11, this matches the actual amount in the dataset. in Kings3 and kings4 the number of collums read is 1 since the function saparetes values using ;
4. Show the dataset so that we can see how R interprets each column¢

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv("data/kings.csv")

class(kings)


ncol(kings)


glimpse(kings)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r}
# YOUR CODE
kings2 <- kings2 %>%
  mutate(duration = as.numeric(End_year) - as.numeric(Start_year)) %>% 
  filter(!is.na(duration))
```
## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
# YOUR CODE
# Using the pipe %>% to summarize the data
kings_summary <- kings2 %>% 
  summarize(avg_reign = mean(duration, na.rm = TRUE))

# Show the result
kings_summary
#The average duration of reign for all rulers is approximately 19.67 years.

## How many and which kings enjoyed a longer-than-average duration of reign?

#You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

# YOUR CODE
long_ruling_kings <- kings2 %>% 
  filter(duration > 19.67)

# Display the result (to see 'which' kings)
long_ruling_kings

# Count the resulting rows (to see 'how many')
long_ruling_kings %>% 
  count()
#there was 25 kings who reigned longer than the average. 
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE
# Sort, Slice, and Calculate
top_three_kings <- kings2 %>% 
  arrange(desc(duration)) %>%        # Sort: Longest reigns at the top
  slice(1:3) %>%                     # Select: Rows 1, 2, and 3
  mutate(Days = duration * 365.25)   # Calculate: duration times days in a year (plus leap days)

# Display the result
top_three_kings
#the Three longest serving monarchs are Christian d.4 with 21915 days, Margrete d. 2 with 18627 days and Erik 7. af Pommern with 15705 days 
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
# 1. Create the midyear column
# Note: Ensure your column names (startYear/endYear) match your dataset exactly!
kings <- kings2 %>% 
  mutate(
    # Wrap the column names in as.numeric()
    Start_year = as.numeric(Start_year),
    End_year = as.numeric(End_year),
    # Now the math will work
    midyear = Start_year + (End_year - Start_year) / 2)
# 2. Create the plot
ggplot(kings, aes(x = midyear, y = duration)) +
  geom_point() +                          # Adds the individual dots for each king
  geom_smooth(method = "lm", se = FALSE) + # Adds a linear trend line (lm = linear model)
  theme_bw() +                            # BONUS: Black and white theme
  labs(
    title = "Trend of Danish Monarch Reign Durations",
    x = "Year (Mid-point of Reign)",
    y = "Years on Throne",
    caption = "Data source: kings.csv"
  ) +
  theme(text = element_text(size = 14))    # BONUS: Increase font size for legibility

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


